\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-21 tex/paper.tex
update method and theory sections

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_z: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_z$ denotes a dependence
of the functional $X$ on a nuisance parameter $Z$ whose values $z \in {\cal Z}$  define a
parameterized family of its systematic uncertainties. That is, $X_z$ and $Y$
induce together a joint probability distribution $p(X,Y|z)$, where the
conditional on $z$ denotes $X_z$. For training, let further assume a finite set $\{
x_i, y_i, z_i \}_{i=1}^N$ of realizations $X_{z_i}(\omega_i), Y(\omega_i)$, for
$\omega_i \in \Omega$ and known values $z_i$ of the nuisance parameter. Our goal
is to learn a function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal Y}$ of
parameters $\theta_f$ (e.g., a neural network-based classifier if ${\cal Y}$ is
a finite set of classes) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
the cross-entropy). In addition, we require that $f(X_z ; \theta_f)$ should be
robust to the value $z$ of the nuisance parameter  -- which remains unknown at
test time. More specifically, we aim at building $f$ such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{z}(\omega) ; \theta_f) = f(X_{z^\prime}(\omega) ; \theta_f)
\end{equation} for all
samples $\omega \in \Omega$ and all $z, z^\prime$ pairs of values of the
nuisance parameter.

Since we do not have training tuples $(X_{z}(\omega),
X_{z^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{align}\label{eqn:criterion-measure}
    &P(\{ \omega | f(X_{z}(\omega) ; \theta_f) = y \} ) \nonumber \\
    &\indent = P( \{ \omega' | f(X_{z^\prime}(\omega') ; \theta_f) = y\})
\end{align}
for all $y \in {\cal Y}$. In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance. Note that a function
$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = y \}$ and $\{
\omega' | f(X_{z^\prime}(\omega'); \theta_f) = y \}$ do not need to be the same
for the equality to hold. In order to simplify notations, and as only
Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
from here on the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) | z ) = p(f(X ; \theta_f) | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values of $f(X ; \theta_f)$.

@@

Adversarial training was first proposed by \citep{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z \sim p_Z$. More specifically, the authors pit a generative model $g:
\mathbb{R} \mapsto \mathbb{R}^p$ against an adversary classifier $d :
\mathbb{R}^p \mapsto \{ 0, 1\}$ whose antagonistic objective is to recognize
real data $X$ from generated data $g(Z)$. Both models $g$ and $d$ are trained
simultaneously, in such a way that $g$ learns to produce samples that are
difficult to identify by $d$, while $d$ incrementally adapts to changes in $g$.
At the equilibrium, $g$ models a distribution whose samples can be identified by
$d$ only by chance. That is, assuming enough capacity in $d$ and  $g$, the
distribution $p_{g(Z)}$ eventually converges towards the real distribution
$p_X$.

@@

    \caption{Architecture for the adversarial training of a binary classifier
    $f$ against a nuisance parameter $Z$. The adversary $r$ models the
    distribution $p(Z|f(X;\theta_f))$ of the nuisance as observed only through the output $f(X;\theta_f)$ of the classifier. By
    maximizing the antagonistic objective ${\cal L}_r(\theta_f, \theta_r)$ (as part of minimizing ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$), the classifier
    $f$ forces $p(Z|f(X;\theta_f)$ to become uniform, which happens
    when $f(X;\theta_f)$ is  independent of the nuisance parameter $Z$ and therefore pivotal.}

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Figure~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f))$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations of $f(X; \theta_f)$, for the current value $\theta_f$ of $f$
parameters, and produces as output a function $p_{\theta_r}(z | f(X;\theta_f))$
modeling the posterior probability density that $z$ parameterizes the sample $X$
observed through $f(\cdot;\theta_f)$.
Intuitively, if $p(f(X; \theta_f)|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f))$. Similarly, if $Z$ takes continuous values and
if we assume some parameteric distribution ${\cal P}$ for $Z|f(X;\theta_f)$
(e.g., a mixture of gaussians, as modeled with a mixture
density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
represented e.g. as network whose output $j$ is the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). As in
\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
$p_{\theta_r}(z|f(X;\theta_f))$ can then be evaluated for any $z \in {\cal Z}$.
As further explained in the next section, let us note that the adversary $r$ may
take any form, i.e. it does need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f))$ of sufficient capacity
to represent the true distribution within its bounded support.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Section~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{X} \mathbb{E}_{Y|X} [ -\log p_{\theta_f} (Y|X) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{X}  \mathbb{E}_{Z|f(X;\theta_f)} [-\log p_{\theta_r} (Z|f(X;\theta_f))].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

@@

    \label{alg:adversarial-training}
    \begin{algorithmic}[1]
        \For{$t=1$ to $T$}
            \For{$k=1$ to $K$} \Comment{Update $r$}
                \State{Sample minibatch $\{x_m, z_m \}_{m=1}^M$} of size $M$;
                \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
                $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|f(x_m;\theta_f))  ;$$}
            \EndFor
            \State{Sample minibatch $\{x_m, y_m, z_m \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
            \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
            $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|f(x_m;\theta_f))  \right],$$
            \indent where $p_{\theta_f}(y_m|x_m)$ denotes $1(y_m=0)(1-f(x_m;\theta_f)) + 1(y_m=1)f(x_m;\theta_f))$;}
        \EndFor
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this setting, the nuisance parameter $Z$ is considered as a random variable
of bounded support, for which we require the uniform prior $p(z)$ (for $z \in
{\cal Z}$). Importantly, classification of $Y$ with respect to $X$ is therefore
considered in the context where $Z$ is marginalized out, which means that the
classifier minimizing ${\cal L}_f$ is optimal with respect to $Y|X$, but not
necessarily with $Y|X,Z$. Results hold for a nuisance parameter $Z$ taking
either categorical values or continuous values within a bounded support. By
abuse of notation, $H(Z)$ denotes the differential entropy in this latter
case. Finally, propositions below are derived in a non-parametric setting, by
assuming that both $f$ and $r$ have enough capacity.

\begin{proposition}\label{prop:1}
Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
\theta_r)$ for fixed $\theta_f$. If $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$
for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$, then $f$ is a pivotal quantity.
\end{proposition}

\begin{proof}
Since we assume enough capacity for $r$, the optimal parameters $$\smash{\hat\theta}_r =
\arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
\theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
and therefore $p(z|f(X;\theta_f)) = p(z)$.
Using the Bayes' rule, we write
\begin{align*}
    p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
                       &= p(f(X;\theta_f)),
\end{align*}
which holds for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ and implies that $f$ is a pivotal quantity.
\end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = \mathbb{E}_{X} [H({Y|X}) - H(Z)]$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r)$, in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{X} [ H({Z|f(X;\theta_f)}) ]$ of the conditional distribution of the nuisance. The
value function $E$ can therefore be rewritten as $$E'(\theta_f) = {\cal
L}_f(\theta_f) - \mathbb{E}_{X} [ H({Z|f(X;\theta_f)}) ].$$  In particular, we have the lower
bound $$\mathbb{E}_{X} [ H({Y|X}) - H(Z)] \leq {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $\mathbb{E}_{X} [H({Y|X})]$,

    \item $\smash{\hat\theta}_f$ maximizes the expected entropy
    $\mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$, which happens when all outcomes of
    $Z|f(X;\smash{\hat\theta}_f)$ are equally likely. In other words,
    $p(z|f(X;\smash{\hat\theta}_f)) = p(z)$ for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ since we
    require a uniform prior by construction. Note that in the continuous case,
    the supremum of the differential entropy over continuous distributions on
    the same bounded support is also realized by the uniform distribution over
    that support.
\end{itemize}
When the lower bound is active, we have $p_{\theta_r}(z|f(X;\theta_f)) = p(z)$
because of the second condition, which induces that the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity as a consequence of Proposition~\ref{prop:1}.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) - \mathbb{E}_X [ H({Z|f(X;\theta_f)}) ]$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $\mathbb{E}_{X} [ H({Y|X}) - H(Z)] < {\cal L}_f(\theta_f) - \mathbb{E}_{X} [H({Z|f(X;\theta_f)})]$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

Interestingly, let us emphasize that these results hold using only the (1D) output
of $f$ (in the case of binary classification) as input to the adversary. We
could similarly enforce an intermediate representation of the data to be
pivotal, e.g. as in \citep{ganin2014unsupervised}, but this is in fact not
necessary.

>>>>

Let assume a probability space $(\Omega, {\cal F}, P)$, where $\Omega$ is a
sample space, ${\cal F}$ is a set of events and $P$ is a probability measure.
Let consider the multivariate random variables $X_z: \Omega \mapsto
\mathbb{R}^p$ and $Y: \Omega \mapsto {\cal Y}$, where $X_z$ denotes a dependence
of the functional $X$ on a nuisance parameter $Z$ whose values $z \in {\cal Z}$  define a
parameterized family of its systematic uncertainties. That is, $X_z$ and $Y$
induce together a joint probability distribution $p(X,Y|Z=z)$, where the
conditional denotes $X_z$. For training, let further assume a finite set $\{
x_i, y_i, z_i \}_{i=1}^N$ of realizations $X_{z_i}(\omega_i), Y(\omega_i)$, for
$\omega_i \in \Omega$ and known values $z_i$ of the nuisance parameter. Our goal
is to learn a score function $f(\cdot;\theta_f) : \mathbb{R}^p \mapsto {\cal S} \subseteq {\cal Y}$ of
parameters $\theta_f$ (e.g., a neural network-based probabilistic classifier) and minimizing  a loss ${\cal L}_f(\theta_f)$ (e.g.,
the cross-entropy). In addition, we require that $f(X_z ; \theta_f)$ should be
robust to the value $z$ of the nuisance parameter  -- which remains unknown at
test time. More specifically, we aim at building $f$ such that in the ideal case
\begin{equation}\label{eqn:criterion-true}
f(X_{z}(\omega) ; \theta_f) = f(X_{z^\prime}(\omega) ; \theta_f)
\end{equation} for all
samples $\omega \in \Omega$ and all $z, z^\prime$ pairs of values of the
nuisance parameter.

Since in general we do not have training tuples $(X_{z}(\omega),
X_{z^\prime}(\omega))$ (for the same unknown $\omega$), we propose instead to
solve the closely related problem of finding a predictive function $f$ such that
\begin{align}\label{eqn:criterion-measure}
    &P(\{ \omega | f(X_{z}(\omega) ; \theta_f) = s \} ) \nonumber \\
    &\indent = P( \{ \omega' | f(X_{z^\prime}(\omega') ; \theta_f) = s\})
\end{align}
for all $s \in {\cal S}$. In words, we are looking for a predictive function $f$
which is a pivotal quantity \citep{degroot1986probability} with respect to the
nuisance parameter. That is, such that  the distribution of $f(X_z; \theta_f)$
is invariant with respect to the value $z$ of the nuisance. Note that a function
$f$ for which Eqn.~\ref{eqn:criterion-true} is true necessarily satisfies
Eqn.~\ref{eqn:criterion-measure}. In general, the converse is however not true,
since the sets of samples $\{ \omega | f(X_{z}(\omega); \theta_f) = s \}$ and $\{
\omega' | f(X_{z^\prime}(\omega'); \theta_f) = s \}$ do not need to be the same
for the equality to hold. In order to simplify notations, and as only
Eqn.~\ref{eqn:criterion-measure} is of direct interest in this work, we denote
from here on the pivotal quantity criterion as
\begin{equation}\label{eqn:criterion}
    p(f(X ; \theta_f) = s | z ) = p(f(X ; \theta_f) = s | z^\prime )
\end{equation}
for all $z,z^\prime \in  {\cal Z}$ and all values $s$ of $f(X ; \theta_f)$.

@@

Adversarial training was first proposed by \citep{goodfellow2014generative} as a
way to build a generative model capable of producing samples from random noise
$z$. More specifically, the authors pit a generative model $g:
\mathbb{R} \mapsto \mathbb{R}^p$ against an adversary classifier $d :
\mathbb{R}^p \mapsto [0, 1]$ whose antagonistic objective is to recognize
real data $X$ from generated data $g(Z)$. Both models $g$ and $d$ are trained
simultaneously, in such a way that $g$ learns to produce samples that are
difficult to identify by $d$, while $d$ incrementally adapts to changes in $g$.
At the equilibrium, $g$ models a distribution whose samples can be identified by
$d$ only by chance. That is, assuming enough capacity in $d$ and  $g$, the
distribution of $g(Z)$ eventually converges towards the real distribution
of $X$.

@@

    \caption{Architecture for the adversarial training of a binary classifier
    $f$ against a nuisance parameter $Z$. The adversary $r$ models the
    distribution $p(z|f(X;\theta_f)=s)$ of the nuisance as observed only through the output $f(X;\theta_f)$ of the classifier. By
    maximizing the antagonistic objective ${\cal L}_r(\theta_f, \theta_r)$ (as part of minimizing ${\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r)$), the classifier
    $f$ forces $p(z|f(X;\theta_f)=s)$ towards the prior $p(z)$, which happens
    when $f(X;\theta_f)$ is  independent of the nuisance parameter $Z$ and therefore pivotal.}

@@

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Figure~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$, for the current value $\theta_f$ of $f$
parameters, and produces as output a function $p_{\theta_r}(z | f(X;\theta_f)=s)$
modeling the posterior probability density that $z$ parameterizes the sample
observed as $s$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

If $Z$ takes discrete values, then $p_{\theta_r}$ can be represented e.g. as a
probabilistic classifier $\mathbb{R} \mapsto \mathbb{R}^{|{\cal Z|}}$ whose
output $j$ (for $j=1, \dots, |{\cal Z}|$) is the estimated probability mass
$p_{\theta_r}(z_j|f(X;\theta_f)=s)$. Similarly, if $Z$ takes continuous values and
if we assume some parameteric distribution ${\cal P}$ for $Z|f(X;\theta_f)=s$
(e.g., a mixture of gaussians, as modeled with a mixture
density network~\citep{bishop1994mixture}), then $p_{\theta_r}$ can be
represented e.g. as network whose output $j$ is the estimated value of the
corresponding parameter $\gamma_j$ of that distribution (e.g., the mean,
variance and mixing coefficients of its components). As in
\citep{nix1994estimating,bishop1994mixture}, the estimated probability density
$p_{\theta_r}(z|f(X;\theta_f)=s)$ can then be evaluated for any $z \in {\cal Z}$ and any score $s \in {\cal S}$.
As further explained in the next section, let us note that the adversary $r$ may
take any form, i.e. it does need to be a neural network, as long as it exposes a
differentiable function $p_{\theta_r}(z|f(X;\theta_f)=s)$ of sufficient capacity
to represent the true distribution.

As for generative adversarial networks, we propose to
train $f$ and $r$ simultaneously, which we carry out by considering
the value function
\begin{equation}
    E(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - {\cal L}_r(\theta_f, \theta_r)
\end{equation}
that we optimize by finding the saddle point $(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ such that
\begin{align}
    \smash{\hat\theta_f} &= \arg \min_{\theta_f} E(\theta_f, \smash{\hat\theta_r}), \label{eqn:min_thetaf} \\
    \smash{\hat\theta_r} &= \arg \max_{\theta_r} E(\smash{\hat\theta_f}, \theta_r) \label{eqn:max_thetar}.
\end{align}
Without loss of generality, the adversarial training procedure to obtain
$(\smash{\hat\theta_f}, \smash{\hat\theta_r})$ is formally presented in
Algorithm~\ref{alg:adversarial-training} in the case of a binary classifier $f :
\mathbb{R}^p \mapsto [0,1]$ modeling $p(Y=1|X)$. For reasons further explained
in Section~\ref{sec:theory}, ${\cal L}_f$ and ${\cal L}_r$  are respectively set to the
expected value of the
negative log-likelihood of $Y|X$ under $f$ and of $Z|f(X;\theta_f)$ under
$r$:
\begin{align}
    {\cal L}_f(\theta_f) &= \mathbb{E}_{x \sim X}  \mathbb{E}_{y \sim Y|x} [ -\log p_{\theta_f} (y|x) ], \\
    {\cal L}_r(\theta_f, \theta_r) &= \mathbb{E}_{s \sim f(X;\theta_f)}  \mathbb{E}_{z \sim Z|s} [-\log p_{\theta_r} (z|s)].
\end{align}
The optimization algorithm consists in using stochastic gradient descent
alternatively for solving Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar}.

@@

    \label{alg:adversarial-training}
    \begin{algorithmic}[1]
        \For{$t=1$ to $T$}
            \For{$k=1$ to $K$} \Comment{Update $r$}
                \State{Sample minibatch $\{x_m, z_m, s_m = f(x_m;\theta_f) \}_{m=1}^M$} of size $M$;
                \State{With $\theta_f$ fixed, update $r$ by ascending its stochastic gradient $\nabla_{\theta_r} E(\theta_f, \theta_r) :=$
                $$\nabla_{\theta_r} \sum_{m=1}^M \log p_{\theta_r}(z_m|s_m)  ;$$}
            \EndFor
            \State{Sample minibatch $\{x_m, y_m, z_m, s_m = f(x_m;\theta_f)  \}_{m=1}^M$} of size $M$; \Comment{Update $f$}
            \State{With $\theta_r$ fixed, update $f$ by descending its stochastic gradient $\nabla_{\theta_f} E(\theta_f, \theta_r) :=$
            $$\nabla_{\theta_f}  \sum_{m=1}^M \left[ -\log p_{\theta_f}(y_m|x_m)  +\log p_{\theta_r}(z_m|s_m)  \right],$$
            \indent where $p_{\theta_f}(y_m|x_m)$ denotes $1(y_m=0)(1-s_m) + 1(y_m=1)s_m$;}
        \EndFor
    \end{algorithmic}
    \end{algorithm}
    \end{minipage}
\end{figure*}

@@

In this setting, the nuisance parameter $Z$ is considered as a random variable
of prior $p(z)$ (for $z \in
{\cal Z}$). Importantly, classification of $Y$ with respect to $X$ is therefore
considered in the context where $Z$ is marginalized out, which means that the
classifier minimizing ${\cal L}_f$ is optimal with respect to $Y|X$, but not
necessarily with $Y|X,Z$. Results hold for a nuisance parameter $Z$ taking
either categorical values or continuous values within a bounded support. By
abuse of notation, $H(Z)$ denotes the differential entropy in this latter
case. Finally, the  proposition below is derived in a non-parametric setting, by
assuming that both $f$ and $r$ have enough capacity.

% \begin{proposition}\label{prop:1}
% Let $\hat\theta_r = \arg \max_{\theta_r} E(\theta_f,
% \theta_r)$ for fixed $\theta_f$. If $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$
% for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$, then $f$ is a pivotal quantity.
% \end{proposition}
%
% \begin{proof}
% Since we assume enough capacity for $r$, the optimal parameters $$\smash{\hat\theta}_r =
% \arg \max_{\theta_r} E(\theta_f, \theta_r) = \arg \min_{\theta_r} {\cal L}_r(\theta_f,
% \theta_r)$$ are such that $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z|f(X;\theta_f))$.
% By assumption, $p_{\hat\theta_r}(z|f(X;\theta_f)) = p(z)$,
% and therefore $p(z|f(X;\theta_f)) = p(z)$.
% Using the Bayes' rule, we write
% \begin{align*}
%     p(f(X;\theta_f)|z) &= \frac{ p(z|f(X;\theta_f)) p(f(X;\theta_f)) } { p(z)}\\
%                        &= p(f(X;\theta_f)),
% \end{align*}
% which holds for all $z \in {\cal Z}$ and all values of $f(X;\theta_f)$ and implies that $f$ is a pivotal quantity.
% \end{proof}

\begin{proposition}\label{prop:2}
If there exists a saddle point $(\smash{\hat\theta}_f, \smash{\hat\theta}_r)$
for Eqn.~\ref{eqn:min_thetaf} and \ref{eqn:max_thetar} such that
$E(\hat\theta_f, \hat\theta_r) = H({Y|X}) - H(Z)$, then
$f(\cdot;\smash{\hat\theta}_f)$ is both an optimal classifier and a pivotal
quantity.
\end{proposition}

@@

For fixed $\theta_f$, the adversary $r$ is optimal at $$\hat\theta_r = \arg
\max_{\theta_r} E(\theta_f, \theta_r)  = \arg \min_{\theta_r} {\cal
L}_r(\theta_f, \theta_r),$$ in which case $p_{\hat\theta_r}(z|f(X;\theta_f)) =
p(z|f(X;\theta_f))$ and ${\cal L}_r$ reduces to the expected entropy
$\mathbb{E}_{s \sim f(X;\theta_f)} [ H({Z|f(X;\theta_f)=s}) ]$ of the conditional distribution of the nuisance.
This expectation is nothing else than the conditional entropy of the random variables
$Z$ and $f(X;\theta_f)$ and can be written as $H(Z|f(X;\theta_f))$.
Accordingly, the
value function $E$ can be restated as $$E'(\theta_f) = {\cal
L}_f(\theta_f) -  H({Z|f(X;\theta_f)}).$$  In particular, we have the lower
bound $$H({Y|X}) - H(Z) \leq {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$$
where the equality holds at $\smash{\hat\theta}_f = \arg \min_{\theta_f}
E'(\theta_f)$ only when
\begin{itemize}
    \item $\smash{\hat\theta}_f$ minimizes the negative log-likelihood ${\cal
    L}_f$ of $X|Y$, which happens when $\smash{\hat\theta}_f$ are the parameters
    of an optimal classifier and in which case ${\cal L}_f$ reduces to its
    minimum value $H({Y|X})$,

    \item $\smash{\hat\theta}_f$ maximizes the conditional entropy
    $H({Z|f(X;\theta_f)})$ since $H(Z|f(X;\theta)) \leq H(Z)$.
\end{itemize}
When the lower bound is active, we have $H(Z|f(X;\theta_f)) = H(Z)$
because of the second condition, which happens exactly when $Z$ and $f(X;\theta_f)$
are independent variables. In other words,  the
optimal classifier $f(\cdot;\smash{\hat\theta}_f)$ is also a pivotal
quantity.

@@

Proposition~\ref{prop:2} suggests that if at each step of
Algorithm~\ref{alg:adversarial-training} the adversary $r$ is allowed to reach
its optimum given $f$ (e.g., by setting $K$ sufficiently high) and if $f$ is
updated to improve ${\cal L}_f(\theta_f) -  H({Z|f(X;\theta_f)})$, then $f$
should converge to a classifier which is both optimal and pivotal, provided such
a classifier exists. On many problems of interest though, such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $ H({Y|X}) - H(Z) < {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

Interestingly, let us emphasize that these results hold using only the (1D) output $s$
of $f(\cdot;\theta_f)$ (in the case of binary classification) as input to the adversary. We
could similarly enforce an intermediate representation of the data to be
pivotal, e.g. as in \citep{ganin2014unsupervised}, but this is in fact not
necessary.
